### 2020-07-08 paper.tex
progress on theory

\section{Proof of detailed balance for involution}
Detailed balance is proven by applying a result from~\citet{tierney1998note} for general Metropolis-Hastings moves with deterministic proposals, and using the measure $h(z) := \pi(z) / (\pi(z) + \pi'(z))$ where $\pi'$ 
% See top of page 4, bullet point 2, of Tierney 1998 note

\section{Proof of stationarity for involution}
Stationarity derives from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\section{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various conditions

>>>>

\section{Derivation of pushforward Radon-Nikodym derivative for special case}
TODO: Do some special case where it simplifies to a Jacobian determinant.
Maybe the case where the involution is composed of an involution on discrete part and a diffeomorphism on the continuous part.

\section{Proof of detailed balance for involution}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
$\pi'(A) := \pi(T^{-1}(A))$.
They let $\nu(A) := \pi(A) + \pi'(A)$.
They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
Then, they set:
\[
\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
\]
To use this result, it suffices to show that $h$ simplifes to the following, when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$:
\[
\frac{h(z)}{h(T(z))}= \frac{\pi(z)}{\pi(T(z))} \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
(note that $T$ is our $f$).
What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$.
Then, the density of $\pi$ with respect to $\nu$ is:
\[
h(z) = \frac{\pi(z)}{\nu(z)} = \frac{\pi(z)}{\pi(z) + \pi'(z)}
\]
It suffices to show that:
\[
\frac{h(z)}{h(T(z))} = \frac{\pi(z)}{\pi(T(z))} \frac{\pi(T(z)) + \pi'(T(z))}{\pi(z) + \pi'(z)}
= \frac{\pi(z)}{\pi(T(z))} \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
It suffices to show that:
\[
\frac{\pi(T(z)) + \pi'(T(z))}{\pi(z) + \pi'(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
Let $a(z) = \pi(z) + \pi'(z)$, and note that $a$ is a density with respect to $\mu$.
Then, the ratio above is:
\[
\frac{\pi(T(z)) + \pi'(T(z))}{\pi(z) + \pi'(z)} = \frac{a(T(z))}{a(z)}
\]
Consider the measure $a$ on $Z$ and the pushforward measure $a \circ T^{-1}$ on $Z$.
It suffices to show that
\[ 
1 = \frac{d (a \circ T^{-1})}{d a} = \frac{a(T(z))}{a(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)
\]
$\frac{d (a \circ T^{-1})}{d a} = 1$ follows from the fact that $a$ and $a \circ T^{-1}$ are actually the same measure:
\begin{align}
a(B) &= \pi(B) + \pi'(B) = \pi(B) + \pi(T^{-1}(B))\\
(a \circ T^{-1})(B) &= \pi(T^{-1}(B)) + \pi(T{-1}(T^{-1}(B))) = \pi(T^{-1}(B)) + \pi(B)
\end{align}
Using the chain rule (TODO check this!):
\[
1 = \frac{d (a \circ T^{-1})}{d a}
=   \frac{d (a \circ T^{-1})}{d (\mu \circ T^{-1})}
    \frac{d (\mu \circ T^{-1})}{d \mu}
    \frac{d \mu}{d a}
= a(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{a(z)}
\]
Note that $a$ is actually equivalent to Tierney's $\nu$.
The non-obvious part is the first factor:
\[
\frac{d (a \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d a}{d \mu} (T(z))
\]
TODO: prove me!

\section{Proof of stationarity for involution}
Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\section{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various assumptions that allow us to e.g. rearrange the order of integration, etc.
